### 2012-11-13 thesis/tex/4_visualformalismsforsafetyconstraints.tex
Final version?

This chapter introduces two different approaches of visual formalisms and later states the final desing of the visual language.
One main difficulty in finding a proper visual formalism is to glean the right level between abstraction and expressiveness since this two characteristics appear to be mutually exclusive in some points. A first approach, so called template constraints,
% and its pros and cons 
are described in section~\ref{sec:templateconstraintformalism}. Subsequently a different design - the operator constraint formalism - is presented in section~\ref{sec:operatorconstraintformalism}.
Section~\ref{sec:comparison} gives a comparison of the two formalisms presented before and states which one is used for the prototype.

@@

%It is not a trivial issue to find a
What visual presentation of constraints would suit a healthbot application developer and what are the main types of safety constraints he could want to check?
% TODO: wir haben gedacht, dass es einfach zu benutzen ist
Led by this question, a visual language based on templates has been developed. These templates make different propositions about states and their relations. The user can choose from several graphical constraint types and customize them by parameterising state variables. So far there are two different kinds of template constraints:

@@

	The idea behind this constraint is that propositions about temporal dependencies can be made. Under certain circumstances a visit of a particular state is restricted as long as a specific precondition is fulfilled. An example constraint for this template is: Whenever a person is eating (x), he must brush his teeth (y) before he can go to bed (z). Thus, going to bed without brushing teeth after meal should never happen.
	\item Each visit of state (x) will eventually result in a visit of either state (y) or state (z) [or state (w) [...]]. All of them are actually reachable.

	This constraint type enforces at least one particular event to happen after a specific precondition.
	An example of such a constraint would be: Every time a person enters a supermarket (x) he will eventually decide to buy 
	%a product or not to buy it.
	something (y) or to leave the shop with empty hands (z). He finally has to make a decision, but he can not avoid both possibilities.
	%Nitpicking readers might note 
	Of course the person could also steal something without buying it, and thus not buy anything but leave the shop with full hands. In fact, exactly in this case the constraint would signal an undesirable ``program'' behaviour with its invalidity.
\end{enumerate}

Figure~\ref{fig:templateconstraints} gives a visual suggestion how the two constraint types could look like. Each state of a constraint is symbolized by the robot surface to make the user think he's directly working with the robot. Furthermore the state labels could be replaced by the real screen displays of the regarding states. All parameters of the template constraints shall be directly customizable by the robot surface symbols. Figure~\ref{fig:templateconstraint_dropdown} demonstrates how states could be choosen as parameters.

@@

All template constraints are additionally equipped with textual expressions in order to explain their semantics. It is a template sentence with placeholders for the states available for parameterisation. Whenever there is a change to the constraint, te text is being updated automatically regarding to the parameter states. These additional expressions are also shown in figure~\ref{fig:templateconstraints}.

Each template constraint describes a certain behaviour which might be later used for checks. Thus, specific formulas have to be provided which represent the semantics of the constraints. They have to be in a format so that they can be processed by a model checker. For the first kind of template operator, a linear temporal logic fomula (LTL, see~section~\ref{sec:modelcheckingandltl}) expresses its semantics (with $x$ as ``x is current state'' and similarly for $y$ and $z$):

\begin{equation} \label{eq:template_one}
  \models \Box (x \rightarrow ([(\neg z) \mathcal{U} y] \vee \Box \neg z))
\end{equation}

``Globally is true: Whenever x is the current state, either z is not visited until y or z is never visited''. Compared to the example above it means: Whenever a person eats something (x), he can not go to bed (z) until he brushes his teeth (y). Otherwise, if he never goes to bed, he does not necessarily need to brush his teeth.

The semantics of the second operator template is more complex since it has a secondary condition that all y, z, ... states are actually reachable after a visit of x. Possibilities can not be expressed by LTL, thus, it needs to use computation tree logic (CTL) for describing this auxiliary condition in a second formula. A short explanation about CTL is already given in section~\ref{sec:delbimbo}, but the two new constructs used for this formula are described below.
The main formula for the second template is still using LTL:
\begin{equation} \label{eq:template_two_one}
  \models \Box (x \rightarrow \Diamond (y \vee z (\vee ...)))
\end{equation}
``Globally is true: Whenever x is the current state, in a future step either y or z (or ...) will be visited''. In context of the example mentioned above it means: Every time a person enters a supermarket (x) he will eventually buy something (y) or buy nothing (z). The fact that both states are really potentially reachable is expressed by the following CTL formula. \emph{AG} replaces the $\Box$ sign and stands for ``on all paths is globally true...'', \emph{EF} is related to $\Diamond$ and expresses ``a path exists where is true in future...''.
\begin{equation} \label{eq:template_two_two}
  \models AG (x \rightarrow ((EF y) \wedge (EF z) (\wedge ...)))
\end{equation}
``It is globally true on all pahts: Whenever x is the current state, there is at least one path leading to y and at least one path leading to z''.

@@

The operator constraint formalism is a new approach with a special focus on expressiveness and uniqueness of the visual language.
%For this the abstraction level is adapted in order to 
It is designed as a visual formalism for linear temporal logic having a visual operator for each LTL operator. Since the visual constraints are directly mapped to LTL expressions, the visual language is equal to LTL in expressiveness. If it is considered as easy to use, non-experts can use it as well as experts who already know LTL.

@@

\begin{description}
	\item[(a)] \emph{AND} operator: $\varphi \wedge \psi$.
	\item[(b)] \emph{OR} Operator: $\varphi \vee \psi$.
	\item[(c)] \emph{IF} operator: $\varphi \rightarrow \psi$.\\Instead of an \emph{IMPLIES} operator an \emph{IF} operator is provided. Its meaning seems to be more intuitive for non-experts since it is a common construct in almost every programming language.
	\item[(d)] \emph{NOT} operator: $\neg \varphi$.
	\item[(e)] Proposition: $\rho$.\\Until now there is only the \emph{state proposition} type, which gives evidence about the currently active state. Other types such as numeric or string equations may be added but are less important for the healthcare subject.
\end{description}

In addition to these five logical operators the visual language should be furthermore capable of expressing constraints about future steps, both ``any future state'' (h) and ``the next state'' (g), that ``events should always happen'' (f), and that ``a property must be true until some future event'' (i).

\begin{description}
	\item[(f)] \emph{ALWAYS} operator: $\Box \varphi$.\\$\varphi$ must be true now and in all following states.
	\item[(g)] \emph{NEXT} operator: $\medcircle \varphi$.\\$\varphi$ has to be true in the next state.
	\item[(h)] \emph{FUTURE} operator: $\Diamond \varphi$.\\At least one time - now or in a later state - $\varphi$ must be true.
	\item[(i)] \emph{UNTIL} operator: $[\varphi \mathcal{U} \psi]$.\\Now and in all following states $\varphi$ must be true at least until there is a state with $\psi$ being true. In addition eventually there has to be a future state with $\psi$ being true.
\end{description}

@@

\subsection{Design decisions}

Like the formula representation introduced by Del Bimbo et al.~\cite{520786}, a constraint consists of nested operators and propositions in a hierarchy. However, the new visual formalism surrenders the three dimensional idea and focuses on two dimensional blocks and their compositions instead, which is used by HomeTL~\cite{4341725} as well. Each operator of the visual language is represented by such a block.

The visual language has support for eight unary or binary operator types and one proposition which have their specific semantics. In order to support the reading of constraints, different significant visual appearances of the operator types shall enable easy classification and a faster subconscious identification~\cite{moody-physics-of-notations}. In this case, different colors are used for each operator type.
% aiming for faster perceptibility and understandability. 
A suggestion for the operators' graphical representation is shown in figure~\ref{fig:operators}.

@@

A second effort for making the perceptibility of the visual language as intuitive as possible is done with the orientation of constraints. 
Operators which consist of nested operators have particular meanings for the two dimensions: The ``logical'' vertical read direction and the horizontal direction for variation in time. Thus, logical operators are arranged in a vertical row wereas time relevant operators are aligned horizontally.
Figure~\ref{fig:directions} demonstrates an example constraint composed along the logical and time axes.
\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{directions}
  \caption{Easy understanding of visual constraints due to intuitive read directions.}
  \label{fig:directions}
\end{figure}
It is read from left to right and top to bottom:
``ALWAYS is true: IF state \emph{'Not taken yet'} is active THEN state \emph{'Polling'} can NOT be visited UNTIL state \emph{'Staff notified'} OR state \emph{'Well done!'} is reached.''

The two dimensions form a special requirement also for the layouting of nested operators. Each operator adapts its size so that it surrounds all recursively nested operators. But also the positioning of nested operators within teir parents have to be taken into account.
Whereas the operator's centered positioning and resizing along the logical direction only depends on the sizes of all sub operators, the arrangement within the time axis turns out to be more complex. Here, sub operators can not just be aligned centered as it is the case vertically. To face the determination of horizontal positioning, every operator gets time reference lines which specify its positioning within its parent and the positioning of its own children within itself. As depicted in figure~\ref{fig:referencelines}, logical operators - i.e. \emph{IfThenOperator}, \emph{AndOperator}, \emph{OrOperator}, \emph{NotOperator} and \emph{StateOperator} - have one reference line. Time relevant operators - i.e. \emph{NextOperator}, \emph{FutureOperator}, \emph{AlwaysOperator} and \emph{UntilOperator} - have two such reference lines, one for their own positioning within their parents and one for the positioning of their children.

@@

As postulated in the requirements for the visual language, constraints shall be easy to read. In fact, there is a possibility to improve the presentation of certain visual constraints. Multiple nested operators of the same kind which have no order priorities such as \emph{OR} or \emph{AND} appear graphically unnecessarily complex. These operators can be visually merged in the manner of unifying their border lines, for example. As a result, all associated operators appear as just one operator with a couple of sub operators (see~figure~\ref{fig:nested_and}).

@@

\begin{itemize}	
	\item All editing is performed only by simple drag and drop. Operators and propositions can be moved and dropped onto other operators. A tool bar provides all operator and proposition types to be instantiated and a possibility for deleting existing operators.
	\item In order to simplify the reading of constraints an operator can be hovered over by the mouse pointer. It is intended to be similar to line coloring when reading a digital text.
	\item Every edit to either the constraint or the underlying state machine results in an automatically (re-)validation of the constraint. The result is immediately displayed to the user.
	\item There is no button for constraint creation / editing except operator and proposition providers, which let the developer create new instances. Having no additional buttons reduces complexity and keeps it simple.
\end{itemize}

@@

Due to its arrow based design the visual language for template constraints is a good visualisation of time dependencies. Furthermore, it allows expressing possibilities by the use of CTL formulas which is not possible with the operator costraint formalism.
Nevertheless, the template visualisations have some trouble with clearness of their semantics. Even though there are well defined formulas behind each constraint and textual expressions which explain the meanings of constraints, the visual presentation seems to be ambiguous in some settings what leads to the risk of being interpreted differently. For example, in the context of the visual language it is unclear whether the first constraint template allows state (x) to be visited more than once before state (z) is reached. It might also be problematic to interprete whether more than one of the states $y$, $z$, ... can be actually visited after a visit of $x$ in the second constraint. In contrast, the operator constraint formalism is not ambiguous due to to its direct mapping to LTL wich is consistently defined. All visual operator expressions represent exactly the same semantics as their corresponding textual LTL formulas.

@@

However, there is a second problem with the template constraints.
Whereas the safety constraints specific to the example within the healthcare domain mentioned in section~\ref{sec:healthbotapplication} are expressable by this visual language, there are other common constraints which have no matching template and thus can not be described. For example, the check of a state being visited never or at least once is not possible. For every new constraint type a new template has to be created what means an infinite amount of templates is needed in order to achieve full expressiveness.
On the contrary, the visual language for operator constraints is not limited in expressiveness compared to LTL, disregarding that there is only one proposition type available. But this is an extension issue of the prototype rather than a lack in expressiveness.
% Gefundene probleme, die man erwähnen könnte: beim if operator ist das übergeordnete always nicht intuitiv. der always operator könnte umgedreht sein.

Since a visual language has to be developed which suits non-professionals as well as experts, the risk of ambiguity is undesirable. Especially for a tool for safety and consistency it would form a really bad basis. For this reason, the approach of template constraints is discarded and the operator constraint formallism is used for the editor prototype instead.
Owing to the recursive nesting of operator constraints, a visual formula can be recursively translated to a LTL sentence which can be processed by the prototype using any ordinary model checker.

>>>>

This chapter introduces two different approaches of visual formalisms and states later on the final desing of the visual language.
A main difficulty in finding a proper visual formalism is to glean the right level between abstraction and expressiveness since these two characteristics appear to be mutually exclusive in some points. A first approach, the so called template constraint formalism,
% and its pros and cons 
is described in section~\ref{sec:templateconstraintformalism}. Subsequently a different design - the operator constraint formalism - is presented in section~\ref{sec:operatorconstraintformalism}.
Section~\ref{sec:comparison} gives a comparison of the two formalisms presented before and states which one is suitable to be used for the prototype.

@@

%It is not a trivial issue to find a
Which visual presentation of constraints would suit a healthbot application developer and what are the main types of safety constraints he could want to check?
% TODO: wir haben gedacht, dass es einfach zu benutzen ist
Led by this question, a visual language based on templates has been developed. These templates make different propositions about visits of states and their relations. Application developers can choose from several graphical constraint types and customize them by parameterising state variables. So far there are two different kinds of template constraints:

@@

	In this constraint template, x, y and z are state variables which can be replaced by real states.
	The idea behind this constraint is that propositions about temporal dependencies can be made. Under certain circumstances the visit of a particular state is allowed only if a specific precondition is fulfilled. An example constraint for this template is: Whenever a person is eating (x), he must brush his teeth (y) before he can go to bed (z). Thus, going to bed without brushing teeth after meal should never happen.
	\item Each visit of state (x) will eventually result in a visit of either state (y) or state (z) [or state (w) [...]]. All of them are actually reachable.

	As before, x, y, z and w are state variables.
	This constraint type enforces at least one particular event to happen after a specific precondition.
	An example of such a constraint would be: Every time a person enters a supermarket (x) he will eventually decide to buy 
	%a product or not to buy it.
	something (y) or to leave the shop with empty hands (z). He finally has to make a decision, but he can not avoid each of both possibilities.
	%Nitpicking readers might note 
	Of course the person could also steal something without buying it, and thus not buy anything but leave the shop with full hands. In fact, exactly in this case the constraint would signal an undesirable ``program'' behaviour with its invalidity.
\end{enumerate}

Figure~\ref{fig:templateconstraints} gives a visual suggestion how the two constraint types could look like. Each state of a constraint is visualised by the image of the robot appearance to make the developer think he is directly working with the robot. Furthermore, the state labels could be replaced by the real screen displays of the regarding states.
It has to be noticed that the arrows used for visualizing template constraints are not necessarily equal to transitions of the state machine.
All parameters for the template constraints shall be directly customizable with the robot surface symbols. Figure~\ref{fig:templateconstraint_dropdown} demonstrates how states could be choosen as parameters.

@@

All template constraints are additionally equipped with textual expressions in order to explain their semantics. These textual expressions are template sentences with placeholders for the states available for parameterisation. Whenever there is a change to the constraint, te text is being updated automatically regarding to the parameter states. These additional expressions are also shown in figure~\ref{fig:templateconstraints}.

Each template constraint describes a certain behaviour which might be later used for checks. Thus, specific formulas have to be provided which represent the semantics of the constraints. They have to be in a specially defined format so that they can be processed by a model checker. For the first kind of template operator, a linear temporal logic fomula (LTL, see~section~\ref{sec:modelcheckingandltl}) expresses its semantics (with $x$ as ``x is current state'' and similarly for $y$ and $z$):

\begin{equation} \label{eq:template_one}
  \models \Box (x \rightarrow [(\neg z) \mathcal{W} y])
\end{equation}

``Globally is true: When x is the current state, z can not not be visited before y is visited''. Compared to the example above it means: Whenever a person eats something (x), he can not go to bed (z) before he brushes his teeth (y). Otherwise, if he never goes to bed, he does not necessarily need to brush his teeth.

The semantics of the second operator template is more complex since it has a secondary condition that all y, z, ... states are actually reachable after a visit of x. Possibilities can not be expressed by LTL, thus, it needs to use computation tree logic (CTL) for describing this auxiliary condition in a second formula. A short explanation about CTL is already given in section~\ref{sec:delbimbo}, but the two new constructs used for this formula are described below.
The main formula for the second template uses LTL:
\begin{equation} \label{eq:template_two_one}
  \models \Box (x \rightarrow \Diamond (y \vee z (\vee ...)))
\end{equation}
``Globally is true: When x is the current state, in a future step either y or z (or ...) will be visited''.
%In context of the example mentioned above it means:
Applied to the example mentioned above it means: every time a person enters a supermarket (x) he will eventually buy something (y) or buy nothing (z). The fact that paths really exist in the program which eventually lead to x and y
% both states are really potentially reachable
is expressed by the following CTL formula. \emph{AG} replaces the $\Box$ sign and stands for ``on all paths is globally true...'', \emph{EF} is related to $\Diamond$ and expresses ``a path exists where is true in future...''.
\begin{equation} \label{eq:template_two_two}
  \models AG (x \rightarrow ((EF y) \wedge (EF z) (\wedge ...)))
\end{equation}
``It is globally true on all pahts: Whenever x is the current state, there is at least one path leading to y and at least one path leading to z''.

@@

The operator constraint formalism is a new approach with a special focus on expressiveness and uniqueness of the visual language.
%For this the abstraction level is adapted in order to 
It is designed as a visual formalism for linear temporal logic having a visual operator for each LTL operator. Since the visual constraints are directly mapped to LTL expressions, the visual language is equal to LTL in expressiveness. If it turns out to be usable easily though it is a mathematical concept, it can also be used by non-experts who do not already know LTL.

@@

\begin{description}
	\item[(a)] \emph{AND} (\emph{AndOperator}): $\varphi \wedge \psi$.\\This conjunction evaluates to true only if both $\varphi$ and $\psi$ are true.
	\item[(b)] \emph{OR} (\emph{OrOperator}): $\varphi \vee \psi$.\\At least one $\varphi$ or $\psi$ being true makes this construct true. The result is false when both parameters are false.
	\item[(c)] \emph{IF THEN} (\emph{IfThenOperator}): $\varphi \rightarrow \psi$.\\Instead of an \emph{IMPLIES} operator an \emph{IF} operator is provided. Its meaning seems to be more intuitive for non-experts; moreover it is a common construct in almost every programming language.
	\item[(d)] \emph{NOT} (\emph{NotOperator}): $\neg \varphi$. This operator negates the value of $\varphi$.
	\item[(e)] Proposition (\emph{StateOperator}): $\rho$.\\Until now there is only the \emph{state proposition} type, which gives evidence about the currently active state. Other types such as numeric or string equations may be added but are less important for the healthcare subject.
\end{description}

In addition to these five logical operators the visual language should be furthermore capable of expressing constraints about future steps, both ``any future state'' (h) and ``the next state'' (g), that ``events should always happen'' (f), and that ``a property must be true until some future event'' (i):

\begin{description}
	\item[(f)] \emph{ALWAYS} (\emph{AlwaysOperator}): $\Box \varphi$.\\$\varphi$ must be true now and in all following states.
	\item[(g)] \emph{NEXT} (\emph{NextOperator}): $\medcircle \varphi$.\\$\varphi$ has to be true in the next state.
	\item[(h)] \emph{FUTURE} (\emph{FutureOperator}): $\Diamond \varphi$.\\At least one time - now or in a later state - $\varphi$ must be true.
	\item[(i)] \emph{UNTIL} (\emph{UntilOperator}): $[\varphi \mathcal{U} \psi]$.\\Now and in all following states $\varphi$ must be true at least until there is a state with $\psi$ being true. In addition eventually there has to be a future state with $\psi$ being true.
\end{description}

@@

\subsection{Design decisions}
\label{sec:designdecisions}

Like the formula representation introduced by Del Bimbo et al.~\cite{520786}, a constraint consists of nested operators and propositions in a hierarchy. However, the new visual formalism surrenders the three dimensional idea and focuses on two dimensional graphical blocks and their compositions instead, which is used by HomeTL~\cite{4341725} as well. Each operator of the visual language is represented by such a block.

The visual language has support for eight unary or binary operator types and one proposition which have their specific semantics. In order to support the reading of constraints, different significant visual appearances of the operator types shall enable easy classification and a faster subconscious identification~\cite{moody-physics-of-notations}. For this purpose an individual color is used for each operatot type.
% aiming for faster perceptibility and understandability. 
A suggestion for the operators' graphical representation is shown in figure~\ref{fig:operators}.

@@

A second effort for making the perceptibility of the visual language as intuitive as possible is done with the orientation of constraints. 
Operators which consist of nested operators have particular meanings for the two dimensions: The ``logical'' vertical read direction and the horizontal direction for variation in time. Thus, logical operators are arranged in a vertical row whereas time relevant operators are aligned horizontally.
Figure~\ref{fig:directions} demonstrates an example constraint composed along the logical and time axes.
\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{directions}
  \caption{Easy understanding of visual constraints due to intuitive read directions.}
  \label{fig:directions}
\end{figure}
It is read from left to right and top to bottom:
``ALWAYS is true: IF state \emph{'Not taken yet'} is active THEN state \emph{'Polling'} can NOT be visited UNTIL state \emph{'Staff notified'} OR state \emph{'Well done!'} is reached.''

The two dimensions form a special requirement also for the layouting of nested operators. Each operator adapts its size so that it surrounds all recursively nested operators. But also the positioning of nested operators within teir parents have to be taken into account.
Whereas the operator's centered positioning and resizing along the logical direction only depends on the sizes of all suboperators, the arrangement within the time axis turns out to be more complex. Here, suboperators can not just be aligned centered as it is the case vertically. To face the determination of horizontal positioning, every operator gets time reference lines which specify its positioning within its parent and the positioning of its own children within itself. As depicted in figure~\ref{fig:referencelines}, logical operators - i.e. \emph{IfThenOperator}, \emph{AndOperator}, \emph{OrOperator}, \emph{NotOperator} and \emph{StateOperator} - have one reference line. Time relevant operators - i.e. \emph{NextOperator}, \emph{FutureOperator}, \emph{AlwaysOperator} and \emph{UntilOperator} - have two such reference lines, one for their own positioning within their parents and one for the positioning of their children.

@@

As postulated in the requirements for the visual language, constraints shall be easy to read. In fact, there is a possibility to improve the presentation of certain visual constraints. Multiple nested operators of the same kind which have no order priorities such as \emph{OR} or \emph{AND} appear graphically unnecessarily complex. These operators can be visually merged in the manner of unifying their border lines, for example. As a result, all associated operators appear as just one operator with a couple of suboperators (see~figure~\ref{fig:nested_and}).

@@

\begin{itemize}	
	\item All editing is performed only by simple drag and drop. Operators and propositions can be moved and dropped onto other operators. A tool bar provides all operator and proposition types to be instantiated and a possibility for deleting existing operators.
	\item In order to simplify the reading of constraints an operator can be hovered over by the mouse pointer. It is intended to be similar to line coloring when reading a digital text.
	\item Every individual edit step to either the constraint or the underlying state machine results in an automatically (re-)validation of the constraint. The result is immediately displayed to the user.
	\item There is no button for constraint creation respectively editing except operator and proposition providers, which let the developer create new instances. Having no additional buttons reduces complexity and keeps it simple.
\end{itemize}

@@

Due to its arrow based design the visual language for template constraints is a good visualization of time dependencies. Furthermore, it allows expressing possibilities by the use of CTL formulas which is not possible with the operator costraint formalism.
Nevertheless, the template visualizations have some trouble with clearness of their semantics. Even though there are well defined formulas behind each constraint and textual expressions which explain the meanings of constraints, the visual presentation seems to be ambiguous in some settings which leads to the risk of being interpreted differently. For example, in the context of the visual language it is unclear whether the first constraint template allows state (x) to be visited more than once before state (z) is reached. It might also be problematic to interpret in the second constraint whether more than one of the states $y$, $z$, ... can actually be visited after a visit of $x$. In contrast, the operator constraint formalism is not ambiguous due to to its direct mapping to LTL which is consistently defined. All visual operator expressions represent exactly the same semantics as their corresponding textual LTL formulas.

@@

Additionally, there is a second problem with the template constraints.
Whereas the safety constraints specific to the example within the healthcare domain mentioned in section~\ref{sec:healthbotapplication} are expressable by this visual language, there are other common constraints which have no matching template and thus can not be described. For example, checks of states like ``being visited never'' or ``being visited at least once'' are not possible. For every new constraint type a new template has to be created which means an infinite amount of templates is needed in order to achieve full expressiveness.
On the contrary, the visual language for operator constraints is not limited in expressiveness compared to LTL, disregarding that there is only one proposition type available. But this is an extension issue of the prototype rather than a lack in expressiveness.
% Gefundene probleme, die man erwähnen könnte: beim if operator ist das übergeordnete always nicht intuitiv. der always operator könnte umgedreht sein.

Since a visual language has to be developed which suits non-professionals as well as experts, the risk of ambiguity is undesirable. Especially for a tool for safety and consistency it would form a really bad basis. For this reason, the approach of template constraints is discarded and the operator constraint formalism is used for the editor prototype instead.
Owing to the recursive nesting of operator constraints, a visual formula can be recursively translated to a LTL sentence which can be processed by the prototype using any ordinary model checker.

### 2012-11-13 thesis/tex/5_automatedgenerationofsafetyconstraints.tex
correction update, chapter 4 has a new structure
Almost all correction issues of Andreas and Alwin are realized now (except the ones of Alwin I couldn't read). Chapter 4 and 6 have a new structure.
Final version?

Safety constraints primarily constitute essential properties which describe correct and safe behaviour of safety critical software applications. In order to discover whether a program meets its safety requirements, its behaviour is checked by the constraints. This happens mostly as a last check before software gets released. 
But regarding updating and maintenace work, often further changes are made to the software even after release.
%But even after possible changes to the software in the context of updating and maintenance work,
However, safety constraints can be applied again and assure the adherence of all properties. Safety constraints thus are also suitable for sustaining consistency during the different phases of software life. In fact, not only safety constraints, also properties of program behaviour which are not safety relevant at all are applicable for that. The more constraints are defined the the better can be the reliability.
But as already mentioned before, it might be difficult for developers to find all reasonable %safety
constraints by hand. % which are needed for full safety covery.
In this case it can happen that a prior correct software version becomes faulty over time because of minor adjustments once in a while which are not covered by constraints.
%To help prevent this to happen

As a solution to this problem, the concept of automated constraint generation for state machines is proposed. Once the initial implementation of a program is finished, its behaviour can be analysed. Sgnificant parts of it can be determined automatically and put into constraint representation which are valid on the current program of course.
%constraint suggestions which are valid on the current program can be computed fully automatically. 
On the one hand, the constraint suggestions may help the developer identifying reasonable constraints he did not think about as well as contradictions between program and specification, eg. if a computed constraint does not make sense at all.
On the other hand, all constraints can be revalidated after each change to the software and thus ensure consistency.

In order to support the developer in developing safe programs and to simplify maintainability, automated generation and suggestion of constraints might be helpful. But it's helpful only if the suggested constraints are relevant and reasonable.

@@

Whether suggested constraints are relevant and reasonable
%This however
depends on the domain and use case. It has to be mentioned that the approach of finding an algorithm for constraint generation presented in this work has a special focus on the healtcare use cases and may not be applicable for other domains.
Regarding the important safety constraints postulated in chapter~\ref{chap:goals}, all the relevant states seem to fall into one of the following categories:

@@

%All associated states which match these properties can be combined to a 
In order to compute the required constraints automatically, and possibly additional ones, for each $\alpha$-state all associated $\beta$-states and the one $\gamma$-state have to be determined. Such a group of states is named subgrap from now on. It is defined as follows:
%Each $\alpha$-, the corresponding $\beta$- and several $\gamma$-states form a part of a state set called \emph{subgraph}. In order to retreive valid constraints a subgraph is defined by the following requirements: %as follows

\begin{itemize}
	\item There is a start state ($\alpha$-state) with at least two outgoing transitions.
	\item There is one end state ($\beta$-state) in which all possible paths starting from start state come together again.
	\item All states between the start and end states have no other incoming transitions than the ones originally coming from the start state. That means there is no path to visit these states except through the start state.
	\item The start state may have incoming transitions from states not contained in the subgraph.
	\item The end state may have incoming and outgoing transitions from and to states not contained in the subgraph.
\end{itemize}

@@

\section{Constraint determination}

@@

\begin{equation} \label{eq:first}
  \models \Box (\alpha \Rightarrow \Diamond \beta)
\end{equation}

\begin{equation} \label{eq:second}
  \models \Box (\alpha \Rightarrow \Diamond (\bigvee_{\gamma} \gamma))
\end{equation}

\begin{equation} \label{eq:third}
  \models \Box (\alpha \Rightarrow [\neg \beta \mathcal{U} \bigvee_{\gamma} \gamma])
\end{equation}

@@

\begin{equation} \label{eq:fourth}
  \models \Box (\alpha \Rightarrow [(\bigwedge_{\delta} \neg \delta) \mathcal{U} \beta])
\end{equation}

@@

\begin{equation} \label{eq:seventh}
	\models \Box (\textnormal{'Not taken yet'} \Rightarrow \Diamond (\textnormal{'Well done!'} \vee \textnormal{'Staff notified'})
\end{equation}
%	\item The program will always eventually check the database for new reminder jobs. Thus we avoid that pending reminder are never processed.
%	\item Either a reminded medication is taken properly or caregivers get notified whenever a patient refuses medication intake.

>>>>

Safety constraints primarily constitute essential properties which describe correct and safe behaviour of safety critical software applications. In order to discover whether a program meets certain safety requirements, its behaviour is checked by the constraints. This happens mostly as a last check before software gets released. 
But regarding updating and maintenace work, often further changes are made to the software even after release.
%But even after possible changes to the software in the context of updating and maintenance work,
However, safety constraints can be applied again and assure the adherence of all properties. Safety constraints thus are also suitable for sustaining consistency during the different phases of software life. In fact, not only safety constraints, also properties of program behaviour which are not safety relevant at all are applicable for that. The more constraints are defined the the better can be the reliability.
But as already mentioned before, it might be difficult for developers to find all reasonable %safety
constraints by hand. % which are needed for full safety covery.
In this case it can happen that a prior correct software version becomes faulty over time because of minor adjustments once in a while which are not covered by constraints.
%To help prevent this to happen

As a solution to this problem, the concept of automated constraint generation for state machines is proposed. Once the initial implementation of a program is finished, the behaviour of this state machine can be analyzed. In particular heuristics can be elaborated which determine significant parts of the state machine automatically and put them into constraint representation. These constraints are of course valid on the current program.
On the one hand, the constraint suggestions may help the developer identifying reasonable constraints he did not think about as well as contradictions between program and specification, eg.\ if a computed constraint does not make sense at all.
On the other hand, all constraints can be revalidated after each change to the software and thus ensure consistency.

In order to support the developer in developing safe programs and to simplify maintainability, automated generation and suggestion of constraints for state machines might be helpful. But it is helpful only if the suggested constraints are relevant and reasonable.

@@

Whether suggested constraints are relevant and reasonable
%This however
depends on the domain and use case. It has to be mentioned that the approach of finding an algorithm for constraint generation presented in this work has a special focus on the healtcare use cases and is not necessarily applicable for other domains.
Among others, each safety constraint postulated in chapter~\ref{chap:goals} was analyzed on a state machine program which represents the medication reminder application of the health care project. It turned out that all states which are named in the constraint are always connected in an identical manner and seem to fall into one of the following categories:

@@

%All associated states which match these properties can be combined to a 
In order to compute the required constraints automatically, and possibly additional ones, for each $\alpha$-state all associated $\beta$-states and the one $\gamma$-state have to be determined. Such a group of states of a state machine is named subgraph from now on. It is defined as follows:
%Each $\alpha$-, the corresponding $\beta$- and several $\gamma$-states form a part of a state set called \emph{subgraph}. In order to retreive valid constraints a subgraph is defined by the following requirements: %as follows

\begin{itemize}
	\item There is a start state ($\alpha$-state) with at least two outgoing transitions.
	\item There is one end state ($\beta$-state) which is the first state where all possible paths of the state machine starting from start state come together again.
	\item All states between the start and end state also belong to the subgraph. They have no other incoming transitions than the ones originally coming from the start state. That means there is no path to visit these states except through the start state.
	\item The start state may have incoming transitions from states not contained in the subgraph.
	\item The end state may have self transitions or incoming and outgoing transitions from and to states which are not contained in the subgraph. Transitions back into the subgraph are not allowed except to the start state.
	\item Except self transitions of the end state, no cycles are allowed in the subgraph.
\end{itemize}

@@

\section{Constraint determination}
\label{sec:constraintdetermination}

@@

\begin{equation} \label{eq:first}
  \models \Box (\alpha \rightarrow \Diamond \beta)
\end{equation}

\begin{equation} \label{eq:second}
  \models \Box (\alpha \rightarrow \Diamond (\bigvee_{\gamma} \gamma))
\end{equation}

\begin{equation} \label{eq:third}
  \models \Box (\alpha \rightarrow [\neg \beta \mathcal{U} \bigvee_{\gamma} \gamma])
\end{equation}

@@

\begin{equation} \label{eq:fourth}
  \models \Box (\alpha \rightarrow [(\bigwedge_{\delta} \neg \delta) \mathcal{U} \beta])
\end{equation}

@@

\begin{equation} \label{eq:seventh}
	\models \Box (\textnormal{'Not taken yet'} \rightarrow \Diamond (\textnormal{'Well done!'} \vee \textnormal{'Staff notified'})
\end{equation}
%	\item The program will always eventually check the database for new reminder jobs. Thus we avoid that pending reminder are never processed.
%	\item Either a reminded medication is taken properly or caregivers get notified whenever a patient refuses medication intake.
